---
output:
  word_document: default
  html_document: default
---

# Data cleaning and descriptive analysis of respiratori disease dataset

#1- import all the packages

```{r include=FALSE}
#1. IMPORT THE PACKAGES

library(readxl)
library(dplyr)
library(gtsummary)
library(sdcMicro)
library(tidyverse)
library(lubridate)
library(reshape2)
library(officer)
library(stringi)
library(stringr)
library(stringdist)
library(text2vec)

#creation of fontions 

#creata a fonction

harmonize_characters <- function(var) {
  var <- str_to_lower(var)                      
  var <- str_trim(var)                          
  var <- str_squish(var)                        
  var <- stri_trans_general(var, "Latin-ASCII") 
  var <- str_replace_all(var, "[^a-z0-9 ]", "") 
  return(var)
}
```

#2- IMPORT THE DATASET AND DATA DESCRIPTION

```{r warning=FALSE, include=FALSE}
dataset <- read_excel("C:/Users/ekinz/OneDrive/Desktop/Respiratory diseases database eng_final.xlsx")
```

#3. DATA PREPROCESSING AND CLEANING

##3.1- de-identification

```{r, echo=FALSE}
dataset <- dataset %>%
  select(-c(unique_code,Residence, `Identification number`,Occupation,`Permanent home`,`Immunosuppressive treatment`, `HIV serology during hospitalization`, `Payment method 2 categories`, `HIV status before consultation or hospitalization...106`,`age >= 18 AND years >= 2012 AND groupenosoreel >= 1 AND groupenosoreel ~= 13 AND groupenosoreel ~= 24 AND LieuPEC = 2 AND (becomeh = 1 OR becomeh = 4) (FILTER)`,`Age categories`))

dataset$`previous TB delay (years)`[-1]<-1
dim(dataset)
```

##3.2- -DATA EXPLORATION

```{r include=FALSE}
dataset %>% 
  tbl_summary(by= Sex)
```

#3.3- REPLACE THE NA VALUES BY 999

```{r include=FALSE}


## Recoding dataset$`Internal Medicine Department` into dataset$`Internal Medicine Department_rec`
dataset$`Internal Medicine Department_rec` <- dataset$`Internal Medicine Department`
dataset$`Internal Medicine Department_rec`[dataset$`Internal Medicine Department` == "0"] <- ""
dataset$`Internal Medicine Department_rec`[dataset$`Internal Medicine Department` == "C3"] <- "C3"
dataset$`Internal Medicine Department_rec`[dataset$`Internal Medicine Department` == "C4"] <- "C4"
dataset$`Internal Medicine Department_rec`[is.na(dataset$`Internal Medicine Department`)] <- ""

## Recoding dataset$`Origin service`
dataset$`Origin service` <- dataset$`Origin service` %>%
  fct_recode(
    NULL = "0"
  )

## Recoding dataset$`location of your Tb ant`
dataset$`location of your Tb ant` <- dataset$`location of your Tb ant` %>%
  fct_recode(
    NULL = "0",
    NULL = "3"
  )



  ## Recoding dataset$`History of chronic renal failure`
dataset$`History of chronic renal failure` <- dataset$`History of chronic renal failure` %>%
  fct_recode(
    "No" = "2"
  )

## Recoding dataset$`Type of hepatitis`
dataset$`Type of hepatitis` <- dataset$`Type of hepatitis` %>%
  fct_recode(
    NULL = "0"
  )


## Recoding dataset$`Cancer seat`
dataset$`Cancer seat` <- dataset$`Cancer seat` %>%
  fct_recode(
    NULL = "1",
    "Breast" = "breast",
    "Breast" = "BREAST",
    "Breast" = "breasts",
    "Breast" = "breasts, lungs, liver, lymph nodes",
    "Breast" = "COLD BREAST",
    "Breast" = "left breast",
    "Breast" = "Left breast",
    "Breast" = "LEFT BREAST",
    "Breast" = "LEFT BREAST METASTAT"
  )

## Recoding dataset$`durationHIV (in months)`
dataset$`durationHIV (in months)` <- dataset$`durationHIV (in months)` %>%
  as.character() %>%
  fct_recode(
    "0.2" = "0.200000002980232"
  )

## Recoding dataset$`Preventive cotrimoxazole`
dataset$`Preventive cotrimoxazole` <- dataset$`Preventive cotrimoxazole` %>%
  fct_recode(
    NULL = "0"
  )

## Recoding dataset$ARV
dataset$ARV <- dataset$ARV %>%
  fct_recode(
    "No" = "2"
  )

## Recoding dataset$`ARV adherence`
dataset$`ARV adherence` <- dataset$`ARV adherence` %>%
  fct_recode(
    "bad" = "0"
  )

## Recoding dataset$`Ethyl index`
dataset$`Ethyl index` <- dataset$`Ethyl index` %>%
  as.character() %>%
  fct_recode(
    "0.8" = "0.889999985694885"
  )


## Recoding dataset$`Macroscopy of pleural fluid`
dataset$`Macroscopy of pleural fluid` <- dataset$`Macroscopy of pleural fluid` %>%
  fct_explicit_na("not done")

dataset$`name of final diagnosis`<-dataset$`name of final diagnosis` %>%
   fct_recode(
    NULL = "3"
  )


## Recoding dataset$`clinical form of tuberculosis`
dataset$`clinical form of tuberculosis` <- dataset$`clinical form of tuberculosis` %>%
  fct_recode(
    NULL = "4"
  )

## Recoding dataset$`Diagnostic group`
dataset$`Diagnostic group` <- dataset$`Diagnostic group` %>%
  fct_recode(
    NULL = "27",
    NULL = "28"
  )


## Recoding dataset$`specify diagnosis` into dataset$`specify diagnosis_rec`
dataset$`specify diagnosis` <- dataset$`specify diagnosis` %>%
  fct_recode(
    NULL = "?",
    NULL = "??",
    NULL = "???",
    NULL = "0",
    NULL = "24",
    NULL = "File to see",
    NULL = "file to see??",
    "HIV" = "hiv"
  )

## Recoding dataset$`Secondary diagnosis 1`
dataset$`Secondary diagnosis 1` <- dataset$`Secondary diagnosis 1` %>%
  fct_recode(
    NULL = "0"
  )


## Recoding dataset$`TB location`
dataset$`TB location` <- dataset$`TB location` %>%
  fct_recode(
    NULL = "15",
    NULL = "8"
  )

## Recoding dataset$`Specify the extrapulmonary location`
dataset$`Specify the extrapulmonary location` <- dataset$`Specify the extrapulmonary location` %>%
  fct_recode(
    "Pleural" = "pleura",
    "Pleural" = "pleural",
    "Pleural" = "PLEURAL"
  )


## Recoding dataset$`Complication in hospital`
dataset$`Complication in hospital` <- dataset$`Complication in hospital` %>%
  fct_recode(
    NULL = "1",
    NULL = "31 07 2013"
  )

```

#3.4- replace the NA values of numerical data

```{r warning=FALSE, include=FALSE}
 
dataset$Age[is.na(dataset$Age)]<-mean(dataset$Age, na.rm=TRUE)


###leukocytes
dataset <- dataset %>%
  mutate(`Leukocytes (Per mm3)` = as.numeric(`Leukocytes (Per mm3)`)) %>%
  mutate(`Leukocytes (Per mm3)` = ifelse(is.na(`Leukocytes (Per mm3)`), 
                                         mean(`Leukocytes (Per mm3)`, na.rm = TRUE), 
                                         `Leukocytes (Per mm3)`))


###HHemoglobine Tx
dataset <- dataset %>%
  mutate(`Hemoglobin Tx` = as.numeric(`Hemoglobin Tx`)) %>%
  mutate(`Hemoglobin Tx` = ifelse(is.na(`Hemoglobin Tx`), 
                                         mean(`Hemoglobin Tx`, na.rm = TRUE), 
                                         `Hemoglobin Tx`))


sum(is.na(dataset$`Hemoglobin Tx`))
dataset$`Hemoglobin Tx`<-as.numeric(dataset$`Hemoglobin Tx`)

## replace the NA values of year of consultation

dataset$`Year of consultation`[is.na(dataset$`Year of consultation`)]<-round(mean(dataset$`Year of consultation`, na.rm=TRUE))


### % of lymphocytes in ascites
dataset$`% of lymphocytes in ascites`<- as.numeric(dataset$`% of lymphocytes in ascites`)
str(dataset$ `% of lymphocytes in ascites`)




```

```{r,echo=FALSE}
tbl_summary(dataset,by = Sex)
```

```{r eval=FALSE, include=FALSE}
#harmonize characters
dataset$`type of surgery`<-harmonize_characters(dataset$`type of surgery`)
# Categorize surgeries into 15 groups
dataset <- dataset %>%
  mutate(
    `type of surgery` = case_when(
      is.na(`type of surgery`) ~ "No recent surgery",
      grepl("lung|thoracotomy|lobectomy", `type of surgery`, ignore.case = TRUE) ~ "Thoracic surgery",
      grepl("appendectomy|hernia|abdominal", `type of surgery`, ignore.case = TRUE) ~ "Abdominal surgery",
      grepl("knee|hip|spinal|orthopedic|fusion", `type of surgery`, ignore.case = TRUE) ~ "Orthopedic surgery",
      grepl("heart|coronary|valve", `type of surgery`, ignore.case = TRUE) ~ "Cardiac surgery",
      grepl("liver|cholecystectomy|hepatic", `type of surgery`, ignore.case = TRUE) ~ "Hepatobiliary surgery",
      grepl("c-section|hysterectomy|gynecologic", `type of surgery`, ignore.case = TRUE) ~ "Obstetric/Gynecologic surgery",
      grepl("thyroidectomy|parathyroid|neck", `type of surgery`, ignore.case = TRUE) ~ "Endocrine surgery",
      grepl("breast|lumpectomy|mastectomy", `type of surgery`, ignore.case = TRUE) ~ "Breast surgery",
      grepl("craniotomy|neurosurgery|brain", `type of surgery`, ignore.case = TRUE) ~ "Neurosurgery",
      grepl("prostatectomy|urologic|kidney", `type of surgery`, ignore.case = TRUE) ~ "Urologic surgery",
      grepl("vascular|bypass|angioplasty", `type of surgery`, ignore.case = TRUE) ~ "Vascular surgery",
      grepl("plastic|reconstruction|cosmetic", `type of surgery`, ignore.case = TRUE) ~ "Plastic surgery",
      grepl("ENT|tonsillectomy|sinus", `type of surgery`, ignore.case = TRUE) ~ "ENT surgery",
      grepl("dermatologic|skin|biopsy", `type of surgery`, ignore.case = TRUE) ~ "Dermatologic surgery",
      TRUE ~ "Other"
    )
  )
gtsummary::tbl_summary(dataset, by=Sex)
```

#3.5- clean the categorical variables

```{r eval=FALSE, include=FALSE}
## Standardize and clean the data

dataset <- dataset %>%
  mutate(
    `Cancer seat` = str_to_lower(`Cancer seat`),                      # Convert to lowercase
    `Cancer seat` = str_trim(`Cancer seat`),                          # Remove extra spaces
    `Cancer seat` = str_replace_all(`Cancer seat`, "[^a-z, ]", ""),   # Remove special characters
    `Cancer seat` = ifelse(is.na(`Cancer seat`) | `Cancer seat` == "", "999", `Cancer seat`), # Handle NA or empty values
    `Cancer seat` = case_when(
      `Cancer seat` == "999" ~ "999",  # Keep "999" separate for missing/empty values
      str_detect(`Cancer seat`, "breast") ~ "Breast",       # Groupe : Breast
      str_detect(`Cancer seat`, "lung|bronchi|pulmonary") ~ "Lung", # Groupe : Lung
      str_detect(`Cancer seat`, "bone|marrow") ~ "Bone",   # Groupe : Bone
      str_detect(`Cancer seat`, "skin|kaposi") ~ "Skin",   # Groupe : Skin
      str_detect(`Cancer seat`, "prostate") ~ "Prostate",  # Groupe : Prostate
      str_detect(`Cancer seat`, "ovary|uterus|cervix|reproductive|testis|penis|scrotum") ~ "Reproductive Organs", # Groupe : Organes reproductifs (femmes et hommes)
      TRUE ~ "Others"                                      # Tous les autres (inclut kidney, liver, etc.)
    )
  )


#Summarize the cancer seat
summary_data <- dataset %>%
  group_by(`Cancer seat`) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

## Recoding dataset$`HIV status before consultation or hospitalization...32`
dataset$`HIV status before consultation or hospitalization...32` <- dataset$`HIV status before consultation or hospitalization...32` %>%
  fct_explicit_na("unknown")

##recode the type of surgery
dataset <- dataset %>%
  mutate(
    `type of surgery` = str_to_lower(`type of surgery`),                      # Convert to lowercase
    `type of surgery` = str_trim(`type of surgery`),                           # Remove extra spaces
    `type of surgery` = str_replace_all(`type of surgery`, "[^a-z, ]", ""),    # Remove special characters
    `type of surgery` = ifelse(is.na(`type of surgery`) | `type of surgery` == "", "999", `type of surgery`),  # Handle NA or empty values
    `type of surgery` = case_when(
      `type of surgery` == "999" ~ "999",  
      str_detect(`type of surgery`, "amputation") ~ "Amputation",  # Group: Amputation
      str_detect(`type of surgery`, "fistula|hernia") ~ "Fistula/Hernia Repair",        # Group: Fistula & Hernia
      str_detect(`type of surgery`, "appendectomy") ~ "Appendectomy", # Group: Appendectomy
      str_detect(`type of surgery`, "caesarian|cesarean") ~ "Cesarean Section", # Group: Cesarean Section
      str_detect(`type of surgery`, "cataract") ~ "Cataract Surgery",  # Group: Cataract
      str_detect(`type of surgery`, "colectomy|colostomy") ~ "Colorectal Surgery",      # Group: Colectomy & Colostomy
      str_detect(`type of surgery`, "craniotomy") ~ "Craniotomy",    # Group: Craniotomy
      str_detect(`type of surgery`, "mastectomy") ~ "Mastectomy",    # Group: Mastectomy
      str_detect(`type of surgery`, "prostate|kidney") ~ "Prostate & Kidney Surgery", # Group: Prostate & Kidney
      str_detect(`type of surgery`, "hysterectomy") ~ "Hysterectomy", # Group: Hysterectomy
      str_detect(`type of surgery`, "osteosynthesis|femur|tibia") ~ "Orthopedic Surgery", # Group: Orthopedic
      str_detect(`type of surgery`, "thyroidectomy|parotid|cervical cancer") ~ "Cervical Surgery", # Group: Thyroid & Parotid
      str_detect(`type of surgery`, "laparotomy|nephrectomy") ~ "Laparotomy & Nephrectomy",  # Group: Laparotomy & Nephrectomy
      TRUE ~ "Others"  # All other surgeries
    )
  )

summary_data1 <- dataset %>%
  group_by(`type of surgery`) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))
```

#3.6- data cleaning final diagnosis

```{r eval=FALSE, include=FALSE}

dataset$`name of final diagnosis`<-harmonize_characters(dataset$`name of final diagnosis`)




# Categorization based on keywords
dataset <- dataset %>%
  mutate(
    `name of final diagnosis` = case_when(
      grepl("tuberculosis|tb|tuberculous", `name of final diagnosis`, ignore.case = TRUE) ~ "Tuberculosis",
      grepl("pneumonia|bronchopneumonia|PNEUMOPATHY", `name of final diagnosis`, ignore.case = TRUE) ~ "Pneumonia",
      grepl("kaposi", `name of final diagnosis`, ignore.case = TRUE) ~ "Kaposi",
      grepl("embolism", `name of final diagnosis`, ignore.case = TRUE) ~ "Pulmonary embolism",
      grepl("asthma", `name of final diagnosis`, ignore.case = TRUE) ~ "Asthma",
      grepl("bronchitis|bronchopneumopathy", `name of final diagnosis`, ignore.case = TRUE) ~ "Bronchitis",
      grepl("cancer|adenocarcinoma|metastasis|metastases|tumor|carcinoma|carcinomatous|metastatic|neoplasia",`name of final diagnosis`, ignore.case = TRUE) ~ "Cancer",
      grepl("copd|CHRONIC OBSTRUCTED",`name of final diagnosis`, ignore.case = TRUE) ~ "COPD",
      grepl("sepsis|septic",`name of final diagnosis`, ignore.case = TRUE) ~ "sepsis",
      grepl("COVID 19",`name of final diagnosis`, ignore.case = TRUE) ~ "COVID 19",
      grepl("stroke|failure|embolism",`name of final diagnosis`, ignore.case = TRUE) ~ "AVC and complication",
      TRUE ~ "Other"
    )
  )
```

1.  data cleaning of secondary diagnosis1

```{r eval=FALSE, include=FALSE}
# Define specific diagnostic groups with matching patterns
diagnostic_groups <- list(
  "Systemic Lupus Erythematosus" = c("lupus"),
  "Acute Kidney Injury" = c("acute kidney injury", "AKI", "renal failure"),
  "Diabetes Mellitus" = c("diabetes", "glycemia", "hyperglycemia"),
  "Heart Failure" = c("heart failure", "cardiac failure", "decompensated heart"),
  "Pulmonary Tuberculosis" = c("pulmonary TB", "tuberculosis", "bone TB"),
  "HIV/AIDS" = c("HIV", "AIDS"),
  "Lymphoma" = c("lymphoma", "lymphadenopathy"),
  "Pulmonary Embolism" = c("embolism", "pulmonary embolism"),
  "Cervical Cancer" = c("cervical cancer", "uterine cancer"),
  "Severe Anemia" = c("anemia", "severe anemia"),
  "Peptic Ulcer Disease" = c("ulcer", "gastro", "duodenal"),
  "Chronic Obstructive Pulmonary Disease (COPD)" = c("COPD", "chronic obstructive"),
  "Febrile Gastroenteritis" = c("enteritis", "febrile enteritis"),
  "Ischemic Stroke" = c("stroke", "ischemic", "cerebral infarction"),
  "Cerebral Toxoplasmosis" = c("toxoplasmosis", "cerebral toxoplasmosis"),
  "Peritoneal Carcinomatosis" = c("carcinomatosis", "peritoneal cancer"),
  "Multiple Myeloma" = c("myeloma", "plasma cell disorder"),
  "Respiratory Failure" = c("respiratory failure", "ARDS"),
  "Malignant Neoplasm" = c("cancer", "malignant", "carcinoma"),
  "Other/Uncategorized" = c("complicated", "syndrome")
)

#replace diagnoses
dataset$`Secondary diagnosis 1`<-sapply(dataset$`Secondary diagnosis 1`, function(diagnosis){
  if (is.na(diagnosis)) {
    return(NA) # Preserve NA
  }
  for(group in names(diagnostic_groups)){
    if (any(grepl(paste(diagnostic_groups[[group]],collapse = "|"), diagnosis, ignore.case = TRUE))){return(group)}
  }
  return("other/uncategorized")
})


#regroup secondary diagnosis 2
regroup_diagnostics <- function(diagnosis1) {
  case_when(
    is.na(diagnosis1) ~ "NA",
    str_detect(diagnosis1, regex("kidney|renal|AKI", ignore_case = TRUE)) ~ "Kidney disorders",
    str_detect(diagnosis1, regex("chronic kidney|CKD", ignore_case = TRUE)) ~ "Chronic kidney disease",
    str_detect(diagnosis1, regex("anemia|hemolytic", ignore_case = TRUE)) ~ "Anemia disorders",
    str_detect(diagnosis1, regex("malnutrition|denutrition", ignore_case = TRUE)) ~ "Nutritional disorders",
    str_detect(diagnosis1, regex("diabetes|glycemic", ignore_case = TRUE)) ~ "Diabetes complications",
    str_detect(diagnosis1, regex("infection|sepsis|HIV", ignore_case = TRUE)) ~ "Infectious diseases",
    str_detect(diagnosis1, regex("stroke|ischemic", ignore_case = TRUE)) ~ "Stroke",
    str_detect(diagnosis1, regex("heart|cardio", ignore_case = TRUE)) ~ "Cardiac disorders",
    str_detect(diagnosis1, regex("liver|hepatic", ignore_case = TRUE)) ~ "Liver disorders",
    str_detect(diagnosis1, regex("neurological|encephal", ignore_case = TRUE)) ~ "Neurological disorders",
    str_detect(diagnosis1, regex("cancer|neoplasm|tumor", ignore_case = TRUE)) ~ "Cancer",
    str_detect(diagnosis1, regex("malaria", ignore_case = TRUE)) ~ "Malaria",
    str_detect(diagnosis1, regex("psycho|depression", ignore_case = TRUE)) ~ "Psychological disorders",
    str_detect(diagnosis1, regex("thyroid", ignore_case = TRUE)) ~ "Thyroid disorders",
    str_detect(diagnosis1, regex("gastro|enteritis", ignore_case = TRUE)) ~ "Gastrointestinal disorders",
    str_detect(diagnosis1, regex("urinary|UTI|pyelonephritis", ignore_case = TRUE)) ~ "Urinary disorders",
    str_detect(diagnosis1, regex("skin|kaposi|rash", ignore_case = TRUE)) ~ "Dermatological disorders",
    TRUE ~ "Other"
  )
}

# Application des regroupements
# Apply the regrouping to the column `Secondary diagnosis 2`
dataset$`Secondary diagnosis 2` <- sapply(dataset$`Secondary diagnosis 2`, regroup_diagnostics)




```

#3.5- ASSESSING THE MISSNG VALUES

```{r include=FALSE}
data_long <- melt(is.na(dataset))
tbl_summary(data_long, by=value)

```

```{r include=FALSE}
### compute the perpectages of non missing data for each variables
percent_non_missing <- dataset %>%
  summarise(across(everything(), ~ mean(!is.na(.)) * 100)) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Percent_Non_Missing")

print(percent_non_missing)

list= c()

list<- percent_non_missing$Variable[percent_non_missing$Percent_Non_Missing<10]

list
#dataset<-dataset [,!names(dataset) %in% list ]
```

```{r include=FALSE}
dataset$`Leukocytes (Per mm3)`<-as.numeric(dataset$`Leukocytes (Per mm3)`)
dataset$`Hemoglobin Tx`<-as.numeric(dataset$`Hemoglobin Tx`)
dataset$`% of lymphocytes in ascites`<- as.numeric(dataset$`% of lymphocytes in ascites`)

```

#3.6- analyze of the numerical vatiables

```{r include=FALSE}
##numerical data exploration
numericaldat<- dataset[sapply(dataset, is.numeric)]
#as.data.frame(summary(numericaldat))

# Afficher plusieurs graphiques pour chaque variable
for (var in colnames(numericaldat)) {
  # Organiser trois graphiques côte à côte
  par(mfrow = c(1, 3))  # 1 ligne, 3 colonnes
  
  # Histogramme
  hist(numericaldat[[var]], 
       main = paste("Histogram of", var), 
       xlab = var, 
       col = "skyblue", 
       border = "black")
  
  # Boxplot
  boxplot(numericaldat[[var]], 
          main = paste("Boxplot of", var), 
          xlab = var, 
          col = "lightgreen")
  
  # Calculer les statistiques descriptives
  min_val <- min(numericaldat[[var]], na.rm = TRUE)
  max_val <- max(numericaldat[[var]], na.rm = TRUE)
  mean_val <- mean(numericaldat[[var]], na.rm = TRUE)
  sd_val <- sd(numericaldat[[var]], na.rm = TRUE)
  median_val <- median(numericaldat[[var]], na.rm = TRUE)
  q1 <- quantile(numericaldat[[var]], 0.25, na.rm = TRUE)  # 1er quartile
  q3 <- quantile(numericaldat[[var]], 0.75, na.rm = TRUE)  # 3ème quartile
  sum_mis<- sum(is.na(numericaldat[[var]]))
  # Tableau descriptif (affiché sous forme de texte)
  plot.new()  # Créer un nouveau graphique vide
  title(main = paste("Statistics of", var))  # Titre
  
  # Ajouter les statistiques descriptives comme texte
  text(0.5, 0.8, paste("Mean (SD): ", round(mean_val, 2), " (", round(sd_val, 2), ")", sep = ""), cex = 1.2)
  text(0.5, 0.6, paste("Median (Q1, Q3): ", round(median_val, 2), " (", round(q1, 2), ", ", round(q3, 2), ")", sep = ""), cex = 1.2)
  text(0.5, 0.4, paste("Min: ", round(min_val, 2), sep = ""), cex = 1.2)
  text(0.5, 0.2, paste("Max: ", round(max_val, 2), sep = ""), cex = 1.2)
  text(0.5,0.1, paste("Missing values:", round(sum_mis,2),sep = ""), cex= 1.2)
}

```

### design a summary table of numerical analysis

```{r include=FALSE}
library(gt)

# Function to calculate the number of outliers
count_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  sum(x < lower_bound | x > upper_bound, na.rm = TRUE)
}

# Create the summary table excluding the 'age' variable
summary_table <- dataset %>%
  select(where(is.numeric)) %>% # Select only numeric variables
  summarise(across(
    everything(),
    list(
      Mean = ~ round(mean(., na.rm = TRUE),2),
      SD = ~ round(sd(., na.rm = TRUE),2),
      Min = ~ round(min(., na.rm = TRUE),2),
      Max = ~ round(max(., na.rm = TRUE),2),
      Outliers = ~ round(count_outliers(.),2),
      Missing = ~ round(sum(is.na(.)),2)
    )
  )) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Variable",".value"),
    names_sep = "_"
  ) %>%
  arrange(Missing)

goodtable<-summary_table %>% 
  rownames_to_column("variable") %>% 
  gt() %>% 
  tab_header(title = "SUMMARY ANALYSIS OF NUMERICALS VARIABLES") %>% 
  fmt_number(columns = Mean:Max, decimals = 2) %>% 
  fmt_number(columns = Outliers:Missing, decimals = 0) %>% 
  cols_label(
    Variable="Variable",
    Mean="Mean",
    SD ="STD",
    Min = "Min val",
    Max ="Max val",
    Outliers= "Ourliers",
    Missing= "Missing") %>% 
  tab_style(
    style = list (cell_text(weight = "bold")),
    locations = cells_column_labels(everything())) %>% 
    tab_options(table.font.size = px(14),
                table.border.top.color = "black",
                column_labels.font.weight = "bold",
                table.border.bottom.color = "black",
                data_row.padding = px(5))
                  
    
# View the summary table
goodtable
```

#6. Data cleaning for all variables except some numericals variable

```{r r,echo=FALSE}

dataset <- dataset %>%
  # Step 1: Replace missing values (NA) with the mean, excluding some of them
  mutate(across(where(is.numeric) & !c(`previous TB delay (years)`,`durationHIV (in months)`,`Last CD4 delay in months`,`Last CD4`,`ARV start time (in months)`,`Symptom delay (in days)`,`CD4 count (current or <1 month)`,`clinical form of tuberculosis`,`TB location`,`Smoking index`,`Ethyl index`,`Respiratory rate`,SpO2,`Sedimentation rate`,`Prothrombin rate`,`Pleural proteins (/mm3)`,`Pleural leukocytes (/mm3)`,`ascitic fluid proteins`,`ascites leukocytes`,`% lymphocytes`,`% of lymphocytes in ascites`), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  # Step 2: Identify and replace outliers with Q3, excluding some of them
  mutate(across(where(is.numeric) & !c(`previous TB delay (years)`,`durationHIV (in months)`,`Last CD4 delay in months`,`Last CD4`,`ARV start time (in months)`,`Symptom delay (in days)`,`CD4 count (current or <1 month)`,`clinical form of tuberculosis`,`TB location`,), 
  function(x) {
    Q1 <- quantile(x, 0.25, na.rm = TRUE)
    Q3 <- quantile(x, 0.75, na.rm = TRUE)
    IQR <- Q3 - Q1
    lower_bound <- Q1 - 1.5 * IQR
    upper_bound <- Q3 + 1.5 * IQR
    # Replace outliers with Q3
    ifelse(x < lower_bound, Q1, ifelse(x > upper_bound, Q3, x))
  }))

```

#6.2 - design the summary table after data imputation

```{r,echo=FALSE}
# Create the summary table excluding the 'age' variable
summary_table1 <- dataset %>%
  select(where(is.numeric)) %>% # Select only numeric variables
  summarise(across(
    everything(),
    list(
      Mean = ~ round(mean(., na.rm = TRUE),2),
      SD = ~ round(sd(., na.rm = TRUE),2),
      Min = ~ round(min(., na.rm = TRUE),2),
      Max = ~ round(max(., na.rm = TRUE),2),
      Outliers = ~ round(count_outliers(.),2),
      Missing = ~ round(sum(is.na(.)),2)
    )
  )) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Variable",".value"),
    names_sep = "_"
  ) %>% 
  arrange(Missing)


goodtable1<-summary_table1 %>% 
  rownames_to_column("variable") %>% 
  gt() %>% 
  tab_header(title = "SUMMARY ANALYSIS OF NUMERICALS VARIABLES") %>% 
  fmt_number(columns = Mean:Max, decimals = 2) %>% 
  fmt_number(columns = Outliers:Missing, decimals = 0) %>% 
  cols_label(
    Variable="Variable",
    Mean="Mean",
    SD ="STD",
    Min = "Min val",
    Max ="Max val",
    Outliers= "Ourliers",
    Missing= "Missing") %>% 
  tab_style(
    style = list (cell_text(weight = "bold")),
    locations = cells_column_labels(everything())) %>% 
    tab_options(table.font.size = px(14),
                table.border.top.color = "black",
                column_labels.font.weight = "bold",
                table.border.bottom.color = "black",
                data_row.padding = px(5))
                  
    
# View the summary table
goodtable1
```

#last data cleaning code

```         
```

# save the clean dataset

```{r}
writexl::write_xlsx(dataset,path = "respirdatasetcleangood.xls" )
```

```{r}
writexl::write_xlsx(dataset,path = "respirdatasetcleangood.xls" )
```

descriptive data analysis

```{r echo=FALSE}
# Generate summary table for categorical variables
categorical_vars <- names(dataset)[sapply(dataset, function(x) is.factor(x) || is.character(x))]

gtsummary::tbl_summary(
  dataset,
  by = Sex,
  include = categorical_vars
)


```

#numerical data analysis

```{r echo=FALSE}

##numerical data exploration
numericaldat<- dataset[sapply(dataset, is.numeric)]
# Afficher plusieurs graphiques pour chaque variable
for (var in colnames(numericaldat)) {
  # Organiser trois graphiques côte à côte
  par(mfrow = c(1, 3))  # 1 ligne, 3 colonnes
  
  # Histogramme
  hist(numericaldat[[var]], 
       main = paste("Histogram of", var), 
       xlab = var, 
       col = "skyblue", 
       border = "black")
# Boxplot
  boxplot(numericaldat[[var]], 
          main = paste("Boxplot of", var), 
          xlab = var, 
          col = "lightgreen")
  
  # Calculer les statistiques descriptives
  min_val <- min(numericaldat[[var]], na.rm = TRUE)
  max_val <- max(numericaldat[[var]], na.rm = TRUE)
  mean_val <- mean(numericaldat[[var]], na.rm = TRUE)
  sd_val <- sd(numericaldat[[var]], na.rm = TRUE)
  median_val <- median(numericaldat[[var]], na.rm = TRUE)
  q1 <- quantile(numericaldat[[var]], 0.25, na.rm = TRUE)  # 1er quartile
  q3 <- quantile(numericaldat[[var]], 0.75, na.rm = TRUE)  # 3ème quartile
  sum_mis<- sum(is.na(numericaldat[[var]]))
  # Tableau descriptif (affiché sous forme de texte)
  plot.new()  # Créer un nouveau graphique vide
  title(main = paste("Statistics of", var))  # Titre
  
  # Ajouter les statistiques descriptives comme texte
  text(0.5, 0.8, paste("Mean (SD): ", round(mean_val, 2), " (", round(sd_val, 2), ")", sep = ""), cex = 1.2)
  text(0.5, 0.6, paste("Median (Q1, Q3): ", round(median_val, 2), " (", round(q1, 2), ", ", round(q3, 2), ")", sep = ""), cex = 1.2)
  text(0.5, 0.4, paste("Min: ", round(min_val, 2), sep = ""), cex = 1.2)
  text(0.5, 0.2, paste("Max: ", round(max_val, 2), sep = ""), cex = 1.2)
  text(0.5,0.1, paste("Missing values:", round(sum_mis,2),sep = ""), cex= 1.2)
}

```
